Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2012-08-20 other/confnoncoherent/paper.tex
modified some notation.  It's important that I put the variances etc in the right place.
Added to paper.  Minor code edits.
Added to paper.  Starting the empirical process proofs.  It's been moved to the appendix.
Added more simulations and conclusion.  I think I just need to fill the last Lemma and the two Propositions now.

%opening
\title{Noncoherent least squares estimators of carrier phase and amplitude}
\author{Robby McKilliam$^{1}$, Andre Pollok$^{1}$, Bill Cowley$^{1}$, Vaughan Clarkson$^{2}$ and Barry Quinn$^{3}$ \vspace{0.2cm} \\
\small $^{1}$Institute for Telecommunications Research, University of South Australia, SA, \textsc{Australia} \vspace{0.1cm} \\
\small $^{2}$School of Information Technology \& Electrical Engineering, The University of Queensland, QLD 4067, \textsc{Australia} \vspace{0.1cm} \\
\small $^{3}$Department of Statistics, Macquarie University, Sydney, \textsc{Australia} \\}

\begin{document}

@@

Here we are interested in \emph{noncoherent detection}, where $s_1, \dots, s_L$ are not known at the receiver, and must also be estimated.  This estimation problem has undergone extensive prior study and is often called \emph{multiple symbol differential detection}~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009}.  A practical approach is the least squares estimator,
\begin{equation}\label{eq:hatadef}
\hat{a} = \arg\min_{a \in \complex} \min_{s_1, \dots, s_L \in \calC} \sum_{i = 1}^L \abs{ y_i - a s_i }^2,
\end{equation}
where $\calC$ is the set of symbols from the $M$-PSK constellation.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is white and Gaussian.  As we will show, the estimator can work well even when the noise is not Gaussian.  Mackenthun~\cite{Mackenthun1994} described an algorithm to compute the least squares estimator $\hat{a}$ that requires only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.

@@

\begin{theorem}\label{thm:normality} (Asymptotic normality)
Under the same conditions as Theorem~\ref{thm:consistency}, let $f(r,\phi)$ be the joint pdf of $R_1$ and $\Phi_1$, and let
\[
g(\phi) = \int_{0}^{\infty} r f(r,\phi) dr.
\]
Put $\hat{\lambda}_L = \sfracpart{\theta_0 - \hat{\theta}}$ and $\hat{m}_L = \hat{\rho} - \rho_0 G(0)$. %If $g(\phi)$ is continuous at $\phi = \tfrac{2\pi}{M}k+\tfrac{\pi}{M}$ for each $k = 0, 1, \dots M-1$, then 
Then the distribution of $(\sqrt{L}\hat{\lambda}_L, \sqrt{L}\hat{m}_L)$ converges to the bivariate normal with zero mean and covariance
\[
\left( \begin{array}{cc} 
H^{-2} A & 0 \\
0 & B
\end{array} \right)
\]
as $L \rightarrow \infty$, where
\[
H = G(0) -  2 \sin(\tfrac{\pi}{M}) \sum_{k = 0}^{M-1} g(\tfrac{2\pi}{M}k + \tfrac{\pi}{M}),
\]
\[
A = \expect R_1^2\sin^2\fracpart{\Phi_1}, \;\;\; B = \expect R_1^2 \cos^2\fracpart{\Phi_1}. 
\]
\end{theorem}

@@

We have studied the least squares estimator of carrier phase and amplitude from the observation of $L$ noisy $M$-PSK symbols.  The estimator can be computed in $O(L\log L)$ operations using the algorithm of Mackenthun~\cite{Mackenthun1994}, and is the maximum likelihood estimator in the case that the noise is additive white and Gaussian.  We shows that the phase estimator $\hat{\theta}$ is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator $\hat{\rho}_0$ is biased, and converges to $G(0)\rho_0$.  This bias is large when the signal to noise ratio is small.  It would be interesting to investigate methods for correcting this bias.

>>>>

%opening
\title{Noncoherent least squares estimators of carrier phase and amplitude}
\author{Robby McKilliam$^{1}$, Andre Pollok$^{1}$, Bill Cowley$^{1}$, Vaughan Clarkson$^{2}$ and Barry Quinn$^{3}$ \vspace{0.2cm} \\
\small $^{1}$Institute for Telecommunications Research, University of South Australia, SA, \textsc{Australia} \vspace{0.1cm} \\
\small $^{2}$School of Information Technology \& Electrical Engineering, The University of Queensland, QLD 4067, \textsc{Australia} \vspace{0.1cm} \\
\small $^{3}$Department of Statistics, Macquarie University, Sydney, \textsc{Australia} 
\thanks{Supported under the Australian Government’s Australian Space Research Program.}
}

\IEEEoverridecommandlockouts
\begin{document}

@@

Here we are interested in \emph{noncoherent detection}, where $s_1, \dots, s_L$ are not known at the receiver, and must also be estimated.  This estimation problem has undergone extensive prior study and is often called \emph{multiple symbol differential detection}~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  A practical approach is the least squares estimator,
\begin{equation}\label{eq:hatadef}
\hat{a} = \arg\min_{a \in \complex} \min_{s_1, \dots, s_L \in \calC} \sum_{i = 1}^L \abs{ y_i - a s_i }^2,
\end{equation}
where $\calC$ is the set of symbols from the $M$-PSK constellation.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is white and Gaussian.  As we will show, the estimator can work well even when the noise is not Gaussian.  Mackenthun~\cite{Mackenthun1994} described an algorithm to compute the least squares estimator $\hat{a}$ that requires only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.

@@

\begin{theorem}\label{thm:normality} (Asymptotic normality)
Under the same conditions as Theorem~\ref{thm:consistency}, let $f(r,\phi)$ be the joint pdf of $R_1$ and $\Phi_1$, and let
\[
g(\phi) = \int_{0}^{\infty} r f(r,\phi) dr.
\]
Put $\hat{\lambda}_L = \sfracpart{\theta_0 - \hat{\theta}}$ and $\hat{m}_L = \hat{\rho} - \rho_0 G(0)$. %If $g(\phi)$ is continuous at $\phi = \tfrac{2\pi}{M}k+\tfrac{\pi}{M}$ for each $k = 0, 1, \dots M-1$, then 
Then the distribution of $(\sqrt{L}\hat{\lambda}_L, \sqrt{L}\hat{m}_L)$ converges to the bivariate normal with zero mean and covariance
\[
\left( \begin{array}{cc} 
H^{-2} A & 0 \\
0 & \rho_0^2 B
\end{array} \right)
\]
as $L \rightarrow \infty$, where
\[
H = G(0) -  2 \sin(\tfrac{\pi}{M}) \sum_{k = 0}^{M-1} g(\tfrac{2\pi}{M}k + \tfrac{\pi}{M}),
\]
\[
A = \expect R_1^2\sin^2\fracpart{\Phi_1}, \;\;\; B = \expect R_1^2 \cos^2\fracpart{\Phi_1} - G^2(0). 
\]
\end{theorem}

@@

We have studied the least squares estimator of carrier phase and amplitude from the observation of $L$ noisy $M$-PSK symbols.  The estimator can be computed in $O(L\log L)$ operations using the algorithm of Mackenthun~\cite{Mackenthun1994}, and is the maximum likelihood estimator in the case that the noise is additive white and Gaussian.  We show that the phase estimator $\hat{\theta}$ is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator $\hat{\rho}_0$ is biased, and converges to $G(0)\rho_0$.  This bias is large when the signal to noise ratio is small.  It would be interesting to investigate methods for correcting this bias.

### 2012-08-20 paper1.tex
Added more simulations and conclusion.  I think I just need to fill the last Lemma and the two Propositions now.

%opening
\title{Carrier phase and amplitude estimation for phase shift keying using pilots and data}
\author{Robby McKilliam, Andre Pollok, Bill Cowley  
\thanks{
Supported under the Australian Government’s Australian Space Research Program.
Robby McKilliam, Andre Pollok and Bill Cowley are with the Institute for Telecommunications Research, The University of South Australia, SA, 5095.  Vaughan~Clarkson is with the School of Information Technology \& Electrical Engineering, The University of Queensland, QLD., 4072, Australia.  Barry~Quinn is with the Department of Statistics, Macquarie University, Sydney, NSW, 2109, Australia.
}}

@@

We assume that time offset estimation and matched filtering have been performed and that $L$ noisy $M$-PSK symbols are observed by the receiver.  The received signal is then,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the reciever and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{align*}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \\
&= \sum_{i \in P} \abs{ y_i - a s_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{align*}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is additive white and Gaussian.  However, as we will show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

@@

Let the noise sequence $\{w_i\}$ is complex Gaussian with independent real and imaginary parts having zero mean and variance $\sigma^2$.  The joint density function of the real and imaginary parts is
\[
\frac{1}{2\pi\sigma^2}e^{-(x^2 + y^2)/\sigma^2}
\]
Theorem~\ref{thm:consistency}~and~\ref{thm:normality} hold and, since the distribution of $w_1$ is circularly symetric, the distribution of $R_1e^{j\Phi_1}$ is identical to the distribution of $1 + \frac{1}{\rho_0} w_1$.
%and the joint density function of the real and imaginary parts of $Z$ is
%\[
%p_Z(x,y) = \rho_0^2 p( \rho_0 (x - 1), \rho_0 y ) = \frac{\kappa^2}{2\pi}e^{-\kappa^2(x ^2 - 2x + 1 + y^2) }
%\] 
%where $\kappa = \tfrac{\rho_0}{\sigma}$.  
It can be shown that
\[
g(\phi) = \frac{\cos^2(\phi)}{4\sqrt{\pi}}\left( \frac{e^{-\kappa^2} }{\sqrt{\pi}} + \kappa \Psi(\phi)  e^{-\kappa^2\sin^2(\phi)}\cos(\phi) \right)
\]
where $\kappa = \rho_0/\sigma$ and
\[
\Psi(\phi) = 1 + \erf(\kappa \cos(\phi)) = 1 + \frac{2}{\sqrt{\pi}}\int_0^{\kappa \cos(\phi)} e^{-t^2} dt .
\]
The value of $A_1, A_2, B_1$ and $B_2$ can be efficiently computed by numerical integration using these formula.

@@

\begin{itemize}
\item Run simulations looking at what happens as the number of pilots changes.
\item Simulations showing what happens as $L$ changes for fixed number of pilots.
\item Show that sometimes just using pilots is better!
\end{itemize}

\begin{figure}[tp]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM2-2.mps}
		\caption{Phase error versus SNR for BPSK with $L=4096$.}
		\label{fig:plotphase}
\end{figure}

\begin{figure}[tp]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM4-2.mps}
		\caption{Phase error versus SNR for QPSK with $L=4096$.}
		\label{fig:plotphase}
\end{figure}

\begin{figure}[tp]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM8-2.mps}
		\caption{Phase error versus SNR for $8$-PSK with $L=4096$.}
		\label{fig:plotphase}
\end{figure}

\begin{figure}[tp]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM2-1.mps}
		\caption{Unbiased amplitude error versus SNR for BPSK with $L=4096$.}
		\label{fig:plotphase}
\end{figure}

\begin{figure}[tp]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM4-1.mps}
		\caption{Unbiased amplitude error versus SNR for QPSK with $L=4096$.}
		\label{fig:plotphase}
\end{figure}

\begin{figure}[tp]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM8-1.mps}
		\caption{Unbiased amplitude error versus SNR for $8$-PSK with $L=4096$.}
		\label{fig:plotphase}
\end{figure}

@@

>>>>

%opening
\title{Carrier phase and amplitude estimation for phase shift keying using pilots and data}
\author{Robby McKilliam, Andre Pollok, Bill Cowley, I. Vaughan L. Clarkson and Barry Quinn  
\thanks{
Supported under the Australian Government’s Australian Space Research Program.
Robby McKilliam, Andre Pollok and Bill Cowley are with the Institute for Telecommunications Research, The University of South Australia, SA, 5095.  Vaughan~Clarkson is with the School of Information Technology \& Electrical Engineering, The University of Queensland, QLD., 4072, Australia.  Barry~Quinn is with the Department of Statistics, Macquarie University, Sydney, NSW, 2109, Australia.
}}

@@

We assume that time offset estimation and matched filtering have been performed and that $L$ noisy $M$-PSK symbols are observed by the receiver.  The received signal is then,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the reciever and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{equation}\label{eq:SSdefn}
\begin{split}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a s_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is additive white and Gaussian.  However, as we will show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

@@

Let the noise sequence $\{w_i\}$ be complex Gaussian with independent real and imaginary parts having zero mean and variance $\sigma^2$.  The joint density function of the real and imaginary parts is
\[
\frac{1}{2\pi\sigma^2}e^{-(x^2 + y^2)/\sigma^2}
\]
Theorem~\ref{thm:consistency}~and~\ref{thm:normality} hold and, since the distribution of $w_1$ is circularly symetric, the distribution of $R_1e^{j\Phi_1}$ is identical to the distribution of $1 + \frac{1}{\rho_0} w_1$.
%and the joint density function of the real and imaginary parts of $Z$ is
%\[
%p_Z(x,y) = \rho_0^2 p( \rho_0 (x - 1), \rho_0 y ) = \frac{\kappa^2}{2\pi}e^{-\kappa^2(x ^2 - 2x + 1 + y^2) }
%\] 
%where $\kappa = \tfrac{\rho_0}{\sigma}$.  
It can be shown that
\[
g(\phi) = \frac{\cos^2(\phi)}{4\sqrt{\pi}}\left( \frac{e^{-\kappa^2} }{\sqrt{\pi}} + \kappa \Psi(\phi)  e^{-\kappa^2\sin^2(\phi)}\cos(\phi) \right)
\]
where $\kappa = \rho_0/\sigma$ and
\[
\Psi(\phi) = 1 + \erf(\kappa \cos(\phi)) = 1 + \frac{2}{\sqrt{\pi}}\int_0^{\kappa \cos(\phi)} e^{-t^2} dt .
\]
The value of $A_1, A_2, B_1$ and $B_2$ can be efficiently computed by numerical integration using these formula.

@@

We present the results of Monte-Carlo simulations with the least squares estimator.  In all simulations the noise $w_1,\dots,w_L$ is independent and identically distributed circularly symmetric and Gaussian with real and imaginary parts having variance $\sigma^2$.  Under these conditions the least squares estimator is also the maximum likelihood estimator.  Simulations are run with $M=2,4,8$ (BPSK and QPSK and $8$-PSK) and with signal to noise ratio $\text{SNR} = \tfrac{\rho_0^2}{2\sigma^2}$ between \unit[-20]{dB} and \unit[20]{dB}.  The amplitude $\rho_0=1$ and $\theta_0$ is uniformly distributed on $[-\pi, \pi)$.  For each value of SNR, $T = 5000$ replications are performed to obtain $T$ estimates $\hat{\rho}_1, \dots, \hat{\rho}_T$ and $\hat{\theta}_1, \dots, \hat{\theta}_T$.

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} show the mean square error (MSE) of the phase estimator when $M=2,4,8$ with $L=4096$ and for varying proportions of pilots symbols $\abs{P} = 0, \tfrac{L}{2}, \tfrac{L}{8}, \tfrac{L}{32}, L$.  When $\abs{P} \neq 0$ (i.e. coherent detection) the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}_\pi^2$.  Otherwise, when $\abs{P}=0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}^2$ as in~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  The dots, squares, circles and crosses are the results of Monte-Carlo simulations with the least square estimator.  The solid lines are the MSEs predicted by Theorem~\ref{thm:normality}.  The theorem accurately predicts the behaviour of the phase estimator when $L$ is sufficiently large and when the SNR is not too small.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$ and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  Theorem~\ref{thm:normality} does not model this behaviour.

Figures~\ref{fig:plotampBPSK},~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} show the variance of the amplitude estimator when $M=2,4,8$ and with $L=32, 256, 2048$ and when the number of pilots symbols is $\abs{P} = 0, \tfrac{L}{2}, L$.  The solid lines are the variance predicted by Theorem~\ref{thm:normality}.  The dots and crosses show the results of Monte Carlo simulations.  Each point is computed as $\tfrac{1}{T}\sum_{i=1}^T\big(\hat{\rho}_i - \rho_0G(0)\big)^2$.  This requires $G(0)$ to be known.  In practice $G(0)$ not be known at the reciever, so Figures~\ref{fig:plotampBPSK},~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} serve to validate the correctness of our asymptotic theory, rather than to suggest the practical performance of the estimator.  When SNR is large $G(0)$ is close to $1$ and the bias of the amplitude estimator is small.  However, $G(0)$ grows without bound as the variance of the noise increases, so the bias is significant when SNR is small.

Figure~\ref{fig:plotphaseQPSKmultL} shows the MSE of the phase estimator when $M=4$ and $L=32,256, 2048$ and the number of pilots is $\abs{P}=\tfrac{L}{8},L$.  The figure depicts an interesting phenomenon.  When $L=2048$ and $\abs{P} = \tfrac{L}{8} = 256$ the number of pilots symbols is the same as when $L=\abs{P} = 256$.  When the SNR is small (approximately less than \unit[0]{dB}) the least squares estimator combining the $256$ pilots symbols and also the $2048-256=1792$ data symbols performs \emph{worse} than the estimator that uses only the $256$ pilots symbols.  A similar phenomenon occurs when $L=256$ and $\abs{P} = \tfrac{L}{8} = 32$.  %When the SNR is small it is better to ignore the data symbols and use only the pilot symbols.  
This behaviour suggests modifying the objective function to give the pilots symbols more importance when the SNR is low.  For example, rather the minimise~\eqref{eq:SSdefn} we could instead minimise a weighted version of it,
\[
SS_{\beta}(a, \{d_i, i \in D\}) = \sum_{i \in P} \abs{ y_i - a s_i }^2 + \beta \sum_{i \in D} \abs{ y_i - a d_i }^2,
\]
where the weight $\beta$ would be small when SNR is small and $\beta$ would be near $1$ when SNR is large.  For fixed $\beta$ the asymptotic properties of this weighted estimator could be derived using the techniques we have developed in Sections~\ref{sec:stat-prop-least},~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} and also in~\cite{McKilliam_leastsqPSKpilotsdata_2012appendix}.  One caveat is that the receiver would require knowledge about the noise distribution in order to correctly choose $\beta$.  We do not invesitigate this further here.

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM2-2.mps}
		\caption{Phase error versus SNR for BPSK with $L=4096$.}
		\label{fig:plotphaseBPSK}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM4-2.mps}
		\caption{Phase error versus SNR for QPSK with $L=4096$.}
		\label{fig:plotphaseQPSK}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM8-2.mps}
		\caption{Phase error versus SNR for $8$-PSK with $L=4096$.}
		\label{fig:plotphase8PSK}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM2-1.mps}
		\caption{Unbiased amplitude error versus SNR for BPSK.}
		\label{fig:plotampBPSK}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM4-1.mps}
		\caption{Unbiased amplitude error versus SNR for QPSK.}
		\label{fig:plotampQPSK}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM8-1.mps}
		\caption{Unbiased amplitude error versus SNR for $8$-PSK.}
		\label{fig:plotamp8PSK}
\end{figure}

%\begin{figure}[tp]
%	\centering
%		\includegraphics[width=\linewidth]{code/data/plotM2-3.mps}
%		\caption{Phase error versus SNR for BPSK.}
%		\label{fig:plotphase}
%\end{figure}

\begin{figure}[tp]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM4-3.mps}
		\caption{Phase error versus SNR for QPSK.}
		\label{fig:plotphaseQPSKmultL}
\end{figure}

@@

We considered least squares estimators of carrier phase and amplitude from noisy communications signals that contain both pilot signals, known to the receiver, and data signals, unknown to the receiver.  We focussed on $M$-ary phase shift keying constellations.  The least squares estimator can be computed in $O(L\log L)$ operations using a modification of an algorithm due to Mackenthun~\cite{Mackenthun1994}, and is the maximum likelihood estimator in the case that the noise is additive white and Gaussian.

We showed, under some conditions on the distribution of the noise, that the phase estimator $\hat{\theta}$ is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator $\hat{\rho}_0$ is biased, and converges to $G(0)\rho_0$.  This bias is large when the signal to noise ratio is small.  It would be interesting to investigate methods for correcting this bias.  A method for estimating $G(0)$ at the reciever appears to be required.

Monte Carlo simulations were used to assess the performance of the least squares estimator and also to validate our asymptotic theory.  Interestingly, when the SNR is small, it is counter productive to use the data symbols to estimate the phase (Figure~\ref{fig:plotphaseQPSKmultL}).  This suggests the use of a weigthed objective function.  In order to select the weight it appears that the reciever would need some knowledge about the noise distribution.  It would be interesting to investigate methods for selecting, or estimating, the weight at the reciever.
